---
title: "Final_Memo_(shipments by month)"
output: html_document
---


## Loading in packages & data


```{r}

# Load Tidyverse, janitor and arcos, tidycensus, mapview, ggthemes, scales

library(tidyverse)
library(janitor)
library(tidycensus)

library(arcos)

library(mapview)
library(ggthemes)
library(scales)

```

## Storing key to load in data

```{r}
# store one of our API keys as an object called key
key <- "uO4EK6I"
```

## Loading in data

```{r}

pills_county_month <- read_csv("pills_county_month.csv")
pills_county_month <- clean_names(pills_county_month)

pills_county_month_year <- read_csv("pills_county_month_year.csv")
pills_county_month_year <- clean_names(pills_county_month_year)

pills_state <- read_csv("pills_state.csv")
pills_state <- clean_names(pills_state)

pills_state_month <- read_csv("pills_state_month.csv")
pills_state_month <- clean_names(pills_state_month)

```

# Main Idea: Between December and January, there is a marked dropoff in opioid shipments nationally.

# Evidence: search for Graph #1

 
```{r}

# Graph 1

#The difference from December to January is a drop by a little bit over 788 million. 

pills_state_monthRE <- pills_state_month %>%
  group_by(month) %>%
  summarise(total_pills = sum(total_pills)) 

ggplot(pills_state_monthRE) +
 geom_point(aes(month, total_pills)) +
 labs(x="Month", y="Total pills", title="National total opiod pills by month", subtitle = "All U.S. states total pills by month", caption = "Source: DEA ARCOS database, via Washington Post")
  options(scipen=999)
  
  
```


# Main Idea: In the vast majority of counties, this January dip held true as well. 

# Evidence: search for Table 1 & Chart 1 & 2



```{r}

#create a file with only the opioid shipments for January
pills_county_January <- pills_county_month %>%
  filter(average_annual_population > 10000) %>%
  filter(month =="01") %>%
  select(countyfips, buyer_county, buyer_state, total_pills_jan = total_pills) 
 

#create a file with only the opioid shipments for December 
pills_county_December <- pills_county_month %>%
  filter(average_annual_population > 10000) %>%
  filter(month =="12") %>%
  select(countyfips, buyer_county, buyer_state, total_pills_dec = total_pills) 


# Table 1

#inner join files for January and December by countyfips, county and state while also calculating the difference between the months and the percentage change. 

county_dips <- pills_county_December %>%
  inner_join(pills_county_January, by=c("countyfips","buyer_county", "buyer_state")) %>%
  mutate(pills_difference = total_pills_dec - total_pills_jan) %>%
  mutate(pills_percent_change = ((total_pills_dec - total_pills_jan)/total_pills_dec))



```


# Main Idea: There are a little bit over 2700 counties that follow this January dip trend as compared to a little less than 300 counties that don't follow the trend.

#Evidence: Chart 1 and Chart 2 


```{r}

#Chart 1 for counties not following Janaury dip trend
county_dips_negative <- county_dips %>%
filter(pills_percent_change < 0) 
  
  
county_dips_negative %>%
select(buyer_county,pills_percent_change) %>% 
arrange(pills_percent_change)


```


```{r}

#Chart 2 for counties following January dip trend
county_dips_positive <- county_dips %>%
filter(pills_percent_change > 0) 
  

county_dips_positive %>%
select(buyer_county,pills_percent_change) %>%
arrange(desc(pills_percent_change))

```


# Main Idea: In this handful of counties, we saw some of the biggest dropoffs between December and January in Charlestown and Clark count (based on the pills difference not the percentage change)

# Evidence: search for Charleston Plot and Clark Plot


```{r}

#Chareleston Plot - month by month for this county - a pills difference of over 11 million

Charleston_month_by_month <- pills_county_month%>%
  filter(buyer_county == "CHARLESTON", buyer_state == "SC")

ggplot(Charleston_month_by_month) +
  geom_point(aes(month, total_pills)) +
 labs(x="Month", y="Total pills", title="Charleston County", subtitle = "Opioid pills by month", caption = "Source: DEA ARCOS database, via Washington Post")

options(scipen=999)

```

```{r}

#Clark Plot - month by month for this county - a pills difference of over 9 million


Clark_month_by_month <- pills_county_month %>%
  filter(buyer_county == "CLARK", buyer_state == "NV") 


ggplot(Clark_month_by_month) +
  geom_point(aes(month, total_pills)) +
 labs(x="Month", y="Total pills", title="Clark county", subtitle = "Opioid pills by month", caption = "Source: DEA ARCOS database, via Washington Post")

options(scipen=999)

```



# Main Idea: Here's what happened in terms of pills person in those counties. 

# Evidence: Search for Clark Chart #2 and Charleston Chart #2



```{r}

#Clark Chart #2

Clark_pills_per_person <- pills_county_month %>%
    filter(buyer_county =="CLARK", buyer_state == "NV") %>%
    select (countyfips, month, buyer_county,total_pills, buyer_state, pills_per_person_county_month) %>%
    group_by(month) 
    
    
    

ggplot(Clark_pills_per_person) +
  geom_point(aes(month, pills_per_person_county_month)) +
 labs(x="Month", y="Total pills", title="Clark county", subtitle = "Opoiod pills per person by month", caption = "Source: DEA ARCOS database, via Washington Post")

options(scipen=999)
    

```


```{r}

#Charleston Chart 2

Charleston_pills_per_person <- pills_county_month %>%
    filter(buyer_county =="CHARLESTON", buyer_state == "SC") %>%
    select (countyfips, month, buyer_county,total_pills, buyer_state, pills_per_person_county_month) %>%
    group_by(month) 
    
    
    

ggplot(Charleston_pills_per_person) +
  geom_point(aes(month, pills_per_person_county_month)) +
 labs(x="Month", y="Total pills", title="Charleston county", subtitle = "Opoiod pills per person by month", caption = "Source: DEA ARCOS database, via Washington Post")

options(scipen=999)
    

```
































